Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-04 doc/sc2012/soop.tex
revision

There are two major challenges in implementing an efficient task
scheduler for Legion:
\begin{itemize}
\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in program order.

\item For performance, Legion must hide the extremely long latencies associated
  with machines that have both distributed memory and many levels of
  memory hierarchy.
\end{itemize}
Our scheduler uses three techniques to achieve both correctness and high performance:
\begin{itemize}
\item If it were necessary to compare all pairs of tasks for dependences that 
would be a bottleneck in most Legion programs.  Fortunately,
subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
(see Section~\ref{sec:dep}).

\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
is pipelined and distributed, and also extracts nested parallelism from subtasks.

\item Legion uses a {\em deferred execution model} that decouples the issuing
of operations from when operations are performed.  An issued operation waits for other operations on
which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
to block, allowing Legion to perform other useful work while the operation is pending.

%  For example, Legion may issue a copy
%operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
%an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
%task $a$ completes and task $b$ will not start until the copy completes.

\end{itemize}

Each processor in the machine runs an instance of the SOOP, which handles all requests
for Legion services from tasks executing on the processor.  The SOOP is analagous to the out-of-order
instruction scheduler in superscalar processors, but working at the granularity of tasks and their region
arguments instead of instructions and their register arguments.
Legion's SOOP has a five stage pipeline: dependence analysis (Section~\ref{sec:dep}),
distribution (Section~\ref{sec:dist}),
mapping (Section~\ref{sec:map}),
execution (Section~\ref{sec:exec}),
and clean-up (Section~\ref{sec:clean}).
We discuss each stage in turn.

\subsection{Depedence Analysis}
\label{sec:dep}

When a Legion task runs, it may spawn new subtasks to be executed.  We refer to the
subtasks as {\em children} of the {\em parent} task.  The children of a parent task
are {\em siblings} of each other.  A task always executes on a single processor, but
its subtasks may execute elsewhere.

When a parent task spawns a child task, the child is {\em registered}
with the SOOP on the processor where the parent is executing;
registration records the subtask's logical region arguments and
associated privileges and coherence properties.  Children are
registered in the sequential execution order in which the parent task
spawns them and enqueued for dependence analysis.  For example, in the
circuit simulation in Listing~\ref{lst:code_ex}, the three spawn
statements on lines 32-34 register will register all of the three
kinds of subtasks (in program order) on the processor where {\tt
simulate\_circuit} executes.

If two children have a dependence between their region arguments, the
SOOP will schedule them to run serially to preserve the program's
sequential semantics.  If two child tasks have no dependence then
Legion can schedule them to execute out of order, in parallel.  Thus,
the first stage in the execution of a task is dependence analysis on
the logical regions used by tasks.

\subsubsection{Dependences}
Detecting dependences between a newly registered task $t$ and a previously registered task 
$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
$t'$ that may overlap with a logical region used by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
both regions need only read privileges there is never a dependence, but if either
task needs write or reduction privileges, the coherence modes are compared using
the table in Figure~\ref{fig:dependence}.

\begin{figure}
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}
\caption{Dependence table.}
\label{fig:dependence}
\end{figure}
A {\tt Dep} entry indicates dependence while {\tt None}
indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
A {\tt Cont} entry
indicates a dependence unless a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task
waits for the older task to complete before it begins execution.

The table lists {\em simultaneous} and {\em relaxed} coherence modes
that we have not yet discussed.  Both simultaneous and relaxed
allow other tasks using the region to execute at the same time and differ
only in what data must be observed.  With simultaneous coherence, a task must 
see all updates to the logical region made by other tasks operating on the same region 
simultaneously (i.e., shared memory semantics).  With relaxed coherence, 
a task may or may not observe updates to the logical region by other tasks executing at
the same time.

\subsubsection{Isolation}

@@

The reader may wonder why we only compare sibling tasks for
dependences when presumably arbitrary pairs of tasks may be dependent. 
The following observation shows why dependence analysis can be limited to
sibling tasks.
\begin{observation}
\label{obs:isolation}
\rm
Let $t_1$ and $t_2$ be two sibling tasks with no dependence.  Then no subtask of $t_1$ has a dependence with any subtask
of $t_2$.
\end{observation}
Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
overlap with the regions used by any subtask of $t_2$.  In the case where the regions used by $t_1$ and $t_2$ do overlap
but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.

\subsubsection{Efficient Dependence Analysis}

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
pieces[0].rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).  We reason about the dependences as follows:
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
a non-disjoint, or {\em aliased}, partition.
\item All the remaining pairs of regions used by the two tasks (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not overlap because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, the two subtasks are independent.

A closer look at this example shows that whether there is overlap
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
that the region tree alternates between regions and partitions:
regions are always children of partitions and
partitions are always children of regions.  There
are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region (because $r_1$ and $r_2$ are in
different partitions of the region, as in cases 4 and 5
above), then $r_1$ and $r_2$ may not be disjoint.

Consider subtasks {\tt distribute\_charge(pieces[0],dt)} and
{\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
uses region {\tt pieces[0].rn\_ghost} and the latter uses region {\tt pieces[1].rn\_shr}.
The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
so these two regions are not disjoint.  Since both tasks write their respective subregions,
there is a dependence and the {\tt update\_voltages} task can only run after the
{\tt distribute\_charges} task has completed.

We can now describe the algorithm for the dependence analysis phase of
the SOOP.  The SOOP maintains a local region forest for each task $t$,
the roots of which are $t$'s region arguments and including all
partitions and subregions used within the task.  Dependence analysis
decorates the region tree with the subtasks of $t$, maintaining the
following invariant. Let $r'$ be a region argument to subtask $t'$,
and let $R$ be the set of regions $r''$ such that $r''$ is an argument
to a task $t''$ registered after $t'$ and $t''$ depends on $t'$ because
of overlap between $r''$ and $r'$.  Then $t'$ is listed on the region that
is the least common ancestor of the set $R \cup \{r'\}$.  Moving $t'$ to an
ancestor of $r'$ has the effect of coarsening the dependences involving $t'$
for tasks registered in the future, and this can result in false dependences.
However, the loss of information can be shown to be small and, as we show next, the
benefit is that it dramatically speeds up dependence analysis overall.

We can identify all dependences between sibling tasks and place tasks
in the correct place in the region tree in amortized time ${\cal O}(nd)$,
where $n$ is the total number of region arguments to tasks and $d$ is the depth
of the region tree.  For region $r'$ an argument of task $t'$, the analysis
walks the unique path in the region forest from a root to $r'$.  For each node $m$
on the walk from the root to $r'$ the following actions are taken:
\begin{itemize}
\item Assume $m$ is a region with multiple child partitions, and let $m'$ be the child on the
path to $r'$.  Any task in a subtree of $m$ other than $m'$ on which $t'$ depends 
is moved to $m$.

\item Assume $m$ is a non-disjoint partition, and let $m'$ be the child on the
path to $r'$.  Any task in a subtree of $m$ other than $m'$ that $t'$ depends on is
moved to $m$.

\item Assume $m$ is a disjoint partition, and let $m'$ be the child on the
path to $r'$.  No tasks in a subtree of $m$ other than $m'$ can conflict wtih $t'$; we simply move
on to $m'$.

\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree are moved to $r'$.
\end{itemize}

@@

Thus, there are two parts to depedence analysis of a region argument
$r'$ to a task: the walk from a region root to $r'$ and the off-path work
to move dependent tasks to the least common ancestor on the path.
Clearly the walk to $r'$ is bounded by the depth of the region tree.
The off-path work can be made efficient by maintaining at each node a
record of which subtrees have tasks in them and with what priveleges,
enabling the off-path search to traverse directly and only to
dependent tasks.  The off-path work is proportional to how far a task
is moved back up the region tree.  Thus, for each region argument a
task is initially placed at some depth in the tree and subsequently
only moved up towards the root, so the overall work is (amortized)
${\cal O}(nd)$.  In practice, we have found this level of efficiency,
where dependence analysis is linear in the number of region arguments (and not
the square) and proportional to the depth of the region tree (and not its size)
to be very important in minimizing the overhead of the SOOP.

>>>>

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
is pipelined, distributed and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks.

\item For performance, Legion must hide the extremely long latencies associated
  with machines that have both distributed memory and many levels of
  memory hierarchy.
\end{itemize}

To solve the second problem Legion uses a {\em deferred execution model} that decouples the issuing
of operations from when operations are performed.  An issued operation waits for other operations on
which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
to block.  With deferred execution, it is worthwhile for the SOOP to run (potentially far) ahead of actual execution,
allocating resources to and issuing tasks that must wait because they have data dependences on earlier tasks that have not 
finished executing.   Pipelining and distributing the SOOP further improves throughput and hides latency.

To solve the first problem, detecting and enforcing data dependences
efficiently at run time, the SOOP analyzes, for each task $t$, the
permissions on the regions $t$ uses and ensures that $t$ waits on any
earlier task whose region uses may conflict with $t$'s.  There is a
difficulty, however.  Recall that the programmer's regions are
logical: a region names a set of objects, but a region does not say
where those objects are physically allocated.
Thus, to {\em map} $t$ (assign physical instances for each logical region used by $t$) the SOOP must know
both what tasks $t$ depends on how those other tasks are mapped.

Our SOOP design solves this problem by dividing the analysis of
dependences between two pipeline stages.  The first SOOP stage
computes task dependences at the granularity of logical regions (Section~\ref{sec:dep}).  This
does not give enough information to map $t$, but it does identify the
other tasks that must map before $t$ can map. 
The third SOOP stage
maps $t$ by carrying out a more refined analysis of the (already
mapped) tasks on which $t$ depends (Section~\ref{sec:map}).
Between these stages, the second stage distributes tasks to other processors
to improve locality or load balance the machine (Section~\ref{sec:dist}).
Once the task has been assigned a processor and physical regions it is executed
using deferred execution (Section~\ref{sec:exec}).  The fifth and final SOOP stage
reclaims resources from completed tasks (Section~\ref{sec:clean}).

\subsection{Stage 1: Mapping Depedences}
\label{sec:dep}

When a Legion task runs, it may spawn new subtasks to be executed.  
These subtasks are {\em children} of the {\em parent} task and {\em siblings} of each other.  
A task always executes on a single processor, but
its subtasks may execute elsewhere.

Each processor in the system runs an instance of the SOOP.
When a parent task spawns a child task, the child is {\em registered}
with the SOOP on the parent's processor;
registration records the subtask's logical regions,
privileges and coherence properties.  Children are
registered in the sequential order the parent
spawns them and enqueued for mapping dependence analysis.  In the
circuit simulation in Listing~\ref{lst:code_ex}, the spawn
statements on lines 32-34 register all three
kinds of subtasks (in program order) on the processor where {\tt
simulate\_circuit} executes.

Detecting mapping dependences between a newly registered task $t$ and
a previously registered task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region used by $t'$ that
may overlap with a logical region used by $t$, the privileges and
coherence modes are compared to determine whether a dependence exists.
If both regions need only read privileges there is never a dependence,
but if either task needs write or reduction privileges, the coherence
modes are compared using the table in Figure~\ref{fig:dependence}.
{\tt Dep} indicates dependence while {\tt None}
indicates independence.  {\tt Same} is a dependence unless the two tasks
use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
{\tt Cont} 
indicates a dependence unless a single writer has
atomic coherence.

\begin{figure}
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}
\caption{Dependence table.}
\label{fig:dependence}
\end{figure}

The table lists {\em simultaneous} and {\em relaxed} coherence modes
that we have not yet discussed.  Both modes
allow other tasks using the region to execute at the same time and differ
only in what data must be observed.  With simultaneous coherence, a task must 
see all updates to the logical region made by other tasks operating on the same region 
simultaneously (i.e., shared memory semantics).  With relaxed coherence, 
a task may or may not observe concurrent updates.

@@

A key property of Legion, that dependence analysis is only needed
between sibling tasks:
\begin{observation}
\label{obs:isolation}
\rm
Let $t_1$ and $t_2$ be two sibling tasks with no dependence.  Then no subtask of $t_1$ has a dependence with any subtask
of $t_2$.
\end{observation}
Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
overlap with the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ do overlap
but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
pieces[0].rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).  We reason about the dependences as follows:
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
an aliased partition.
\item All remaining pairs of regions (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not overlap because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, the two subtasks are independent.

A closer look at this example shows that whether there is overlap
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  
There are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region (because $r_1$ and $r_2$ are in
different partitions of the region, as in cases 4 and 5
above), then $r_1$ and $r_2$ may not be disjoint.

Consider subtasks {\tt distribute\_charge(pieces[0],dt)} and
{\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
uses region {\tt pieces[0].rn\_ghost} and the latter uses region {\tt pieces[1].rn\_shr}.
The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
so these two regions are not disjoint.  Since both tasks write their respective subregions,
there is a dependence and the {\tt update\_voltages} task can only map after the
{\tt distribute\_charges} task has mapped.

We can now describe the algorithm for the dependence analysis phase of
the SOOP.  The SOOP maintains a local region forest for each task $t$,
the roots of which are $t$'s region arguments and including all
partitions and subregions used within the task.  Dependence analysis
decorates the region tree with the subtasks of $t$, maintaining the
following invariant. Let $r'$ be a region argument to subtask $t'$,
and let $R$ be the set of regions $r''$ such that $r''$ is an argument
to a task $t''$ registered after $t'$ and $t''$ depends on $t'$
because of overlap between $r''$ and $r'$.  Then $t'$ is listed on the
region that is the least common ancestor of the set $R \cup \{r'\}$.
Moving $t'$ to an ancestor of $r'$ has the effect of coarsening the
dependences involving $t'$ for tasks registered in the future, and
this can result in false dependences.  However, the loss of
information can be shown to be small and only affects when $t$ can
map; any loss of information here will be recovered in the mapping
phase.  Furthermore, the benefit is that it dramatically speeds up
dependence analysis overall.

We can identify all mappoing dependences between sibling tasks and place tasks
in the correct place in the region tree in amortized time ${\cal O}(d)$ per region,
whre $d$ is the depth of the region tree.  For region $r'$ an argument of task $t'$, the analysis
walks the unique path in the region forest from a root to $r'$.  For each node $m$
on the walk from the root to $r'$ the following actions are taken:
\begin{itemize}
\item If $m$ is a region with multiple child partitions and $m'$ is the child on the
path to $r'$, then any task in a subtree of $m$ other than $m'$ on which $t'$ depends 
is moved to $m$.

\item If $m$ is a non-disjoint partition and $m'$ is the child on the
path to $r'$, then any task in a subtree of $m$ other than $m'$ that $t'$ depends on is
moved to $m$.

\item If $m$ is a disjoint partition and $m'$ is the child on the
path to $r'$, then no tasks in a subtree of $m$ other than $m'$ can conflict wtih $t'$; we simply move
on to $m'$.

\item If $m = r'$ then any dependent tasks in $r'$'s subtree are moved to $r'$.
\end{itemize}

@@

There are two parts to this analysis:
the walk from a region root to $r'$ and the off-path work
to move dependent tasks to the least common ancestor on the path.
Clearly the walk to $r'$ is bounded by the depth of the region tree.
The off-path work can be made efficient by maintaining at each node a
record of which subtrees have tasks in them and with what privileges,
enabling the off-path search to traverse directly and only to
dependent tasks.  The off-path work is proportional to how far a task
is moved back up the region tree.  Thus, for each region argument a
task is initially placed at some depth in the tree and subsequently
only moved up towards the root, so the overall work per region is (amortized)
${\cal O}(d)$.  In practice, we have found this level of efficiency,
where analyzing mapping dependences is linear in the number regions used (and not the square) and
proportional to the depth of the region tree (and not its size)
to be very important in minimizing the overhead of the SOOP.
